---
title: "HELIX: data ingestion from the Exposome Data Challenge 2021"
author: "Nicol&oacute; Foppa Pedretti"
date: last-modified
format:
  html:
    fontsize: smaller
    number-sections: true
    embed-resources: true
    toc: false
bibliography: ../references.bib
csl: ../ieee.csl
engine: knitr
---

## Introduction

This document converts the HELIX Exposome Data Challenge 2021 release
[@vrijheid2014helix] from its R formats (`.RData` data frames and Bioconductor
`ExpressionSet` objects [@huber2015bioconductor]) into Arrow IPC files under
`.data/helix/`, the only inputs read by the HELIX analyses of this repository.
It is the one step written in R: `ExpressionSet` is an S4 class that the Julia
`RData` reader cannot convert, while R loads every object natively.

All tables share two conventions, so that the Julia code can join them without
type conversions:

+ every subject identifier is a string column named `ID`;
+ factors are stored as strings, missing values as Arrow nulls.

Omics layers keep the `ExpressionSet` orientation (one row per feature, one
column per sample, first column `feature`) and come with a sample sheet
(`<layer>_cov.arrow`, the `phenoData`) and a feature sheet
(`<layer>_features.arrow`, the `featureData`).

## Setup

The conversion uses `Biobase` to unpack the `ExpressionSet` objects and
`arrow` to write the files.

```{r}
#| label: setup
#| message: false
#| warning: false
library(Biobase)
library(arrow)

# source release (sibling clone of the brge_data_large repository) and target folder
src <- normalizePath(file.path("..", "..", "..", "brge_data_large", "data",
                               "ExposomeDataChallenge2021"), mustWork = TRUE)
out <- file.path("..", "..", ".data", "helix")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
```

Two helpers enforce the conventions above.

```{r}
#| label: helpers
as_plain <- function(df) {
  df[] <- lapply(df, function(v) if (is.factor(v)) as.character(v) else v)
  if ("ID" %in% names(df)) df$ID <- as.character(df$ID)
  df
}

write_table <- function(df, name) {
  path <- file.path(out, paste0(name, ".arrow"))
  write_feather(as_plain(df), path)
  data.frame(file = basename(path), rows = nrow(df), cols = ncol(df))
}

write_eset <- function(eset, name) {
  stopifnot(identical(colnames(exprs(eset)), as.character(pData(eset)$ID)))
  assay <- data.frame(feature = rownames(exprs(eset)), exprs(eset),
                      check.names = FALSE, stringsAsFactors = FALSE)
  feats <- data.frame(feature = rownames(fData(eset)), fData(eset),
                      check.names = FALSE, stringsAsFactors = FALSE)
  rbind(write_table(assay, name),
        write_table(pData(eset), paste0(name, "_cov")),
        write_table(feats, paste0(name, "_features")))
}

load_object <- function(file) {
  e <- new.env()
  nm <- load(file.path(src, file), envir = e)
  mget(nm, envir = e)
}
```

## Exposome, covariates and phenotypes

`exposome.RData` holds four data frames for the 1301 children: 222 exposures
(pregnancy and postnatal), 13 covariates, 6 phenotypes and the codebook that
describes every variable.

```{r}
#| label: exposome-tables
ex <- load_object("exposome.RData")
rbind(write_table(ex$exposome,   "exposome"),
      write_table(ex$covariates, "covariates"),
      write_table(ex$phenotype,  "phenotype"),
      write_table(ex$codebook,   "codebook"))
```

## Omics layers

Urine metabolome (44 metabolites), serum metabolome (177 metabolites),
proteome (36 proteins) and blood transcriptome (28738 transcript clusters).

```{r}
#| label: omics-layers
#| message: false
layers <- c(met_urine = "metabol_urine.Rdata", met_serum = "metabol_serum.Rdata",
            proteome  = "proteome.Rdata",      genexpr   = "genexpr.Rdata")
do.call(rbind, lapply(names(layers), function(name) {
  write_eset(load_object(layers[[name]])[[1]], name)
}))
```

## Output

Every file written to `.data/helix/`, read back to confirm that the
round trip preserves the dimensions.

```{r}
#| label: output-check
files <- list.files(out, pattern = "\\.arrow$", full.names = TRUE)
files <- files[basename(files) != "clean_dt.arrow"]   # built later by 01_data_prep
do.call(rbind, lapply(files, function(f) {
  tb <- read_feather(f, as_data_frame = FALSE)
  data.frame(file = basename(f), rows = tb$num_rows, cols = tb$num_columns,
             MB = round(file.size(f) / 2^20, 1))
}))
```

## References

::: {#refs}
:::
